Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2020-01-14 paper/manuscript.tex
changing afe to alchemical

Alchemical free energy (AFE) calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor or the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}. 
\todo[inline, color={yellow!20}]{JDC: Do we really want to introduce the abbreviation "AFE" when we can just use the word "alchemical" wherever it might be needed?}

@@

Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with conventional computational methods, but are more readily tractable with AFE methods. 

@@

In AFE simulations, the introduction of intermediate \textit{alchemical states} that bridge the high-probability regions of configuration space between the two physical endstates of interest permits the free energy of large transformations to be robustly computed.

@@

AFE calculations can be used in a variety of scenarios, such as computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} (A)); 

@@

AFE calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (D))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (E))~\cite{mobley2007predicting,aldeghi2016accurate,aldeghi2017predictions}; and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (F))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.

Once an AFE simulation as been performed, the data must be analyzed to estimate the free energy for the transformation of interest.

@@

As the field of molecular simulation moves towards millisecond timescales, accurate AFE calculations on even more challenging problems will become reasonable to perform. In the meantime, today's users may find it hard to get started with these complex calculations whilst also keeping up with the fast pace of change. This best practice guide aims to serve as a go-to reference for current recommendations and tips for users of all experience.  

@@

We assume the reader possesses a basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association. If you feel unfamiliar with some of these concepts please good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}. Oftentimes, docking calculations are required to generated initial binding pose for a small molecule for an AFE calculation, where binding poses are not known, thus some basic familiarty on how to run docking calculations is also expected~\cite{grinter2014challenges}. 

@@

In the remainder of this paper, we will cover topics that are key to the preparation~(\ref{sec:step1}), choice and use of correct protocols~(\ref{sec:step2}), and finally the best practices that should be used for the analysis of AFE calculations~(\ref{sec:step3}). 

@@

    \textit{AFE} - Alchemical Free Energy\\

@@

Many practitioners expect AFE methods to provide valuable guidance for drug discovery, and to exhibit accuracy superior to most  alternative approaches for suitable targets~\cite{kuhn2017prospective}. It has been stressed that successful application in industry may require considerable knowledge of the "domain of applicability" of free energy calculations -- where they work well and where they will not ~\cite{sherborne2016collaborating}. Successful application also requires robust protocols for preparing, submitting and analysing AFE calculations. In this regard, the issues mentioned in the previous section such as understanding the suitability and timescales to capture the structure activity relationships (SAR), and performing up-front tests of performance are all relevant to drug discovery applications. Without venturing too far into details of system setup, which is beyond the scope of this article, we highlight some critical factors affecting accuracy and successful application. 

@@

    \caption{\textbf{Two common topologies for AFE calculations} left: A single topology uses softcore potentials to convert from one type of atom to an other. Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms. right: The dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses softcore potentials for this. The 'mixed' intermediate atoms are used in both dual and single topology approaches. Only the way the transformation occurs and the end states differ. Following the arrow along the left and right illustrate the differences.  Figure adapted from \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}}

@@

\subsubsection{What different sampling schemes exist and which one should I use?}
With all AFE simulations having the requirement of sampling from different sets of $\lambda$ states in common different approaches have been designed. Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes used. The simplest approach is defining a set of fixed $\lambda$ values and then run independent simulations at each of the intermediate $\lambda$ values, see Fig.~\ref{fig:fig_sampling_scheme}(A). This type of scheme is currently used for AMBER TI calculations~\cite{song2019} and for BioSimSpace~\cite{hedges2019biosimspace}. A slightly more complex approach is allowing replica mixing, meaning that each independent $\lambda$ simulation is attempted to be exchanged with a neighbouring $\lambda$ replica. This is based on ideas of Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{wendsen1986replica}. Using a Metropolis Hastings based acceptance criterion for exchanges means that the generated ensemble of all replicas is still sampling from the Boltzmann distribution and has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensionala,sugita1999replicaexchangea, woods2003developmenta, jiang2010free}. The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}(B). It is supported in various software packages that provide AFE implementations, such as GROMACS~\cite{aldeghi2015accurate}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}. A third approach borrows ideas from simulated tempering~\cite{marinari1992simulateda}. In this scheme a single replica rapidly explores all of $\lambda$ space by optimally working out weights that allow switching between different intermediate $\lambda$ values, as seen in Fig.~\ref{fig:fig_sampling_scheme}(C) and also referred to as self adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in~\cite{andrearizzi2019choderalab}. The last approach available in conjunction with GROMACS makes use of non-equilibrium simulations~\cite{aldeghi2018accurate}. In this approach, only end state $\lambda$ replicas ($\lambda$=0, $\lambda=1$) are simulated at equilibrium intermediate information is generated from non-equilibrium simulations. A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme}. 
\ref{fig:fig_sampling_scheme} A:\cite{hedges2019biosimspace, song2019using } B\cite{aldeghi2015accurate, wang2015accurate,jiang2019computing} C:\cite{li2007simulated, tan2017optimally} D:\cite{aldeghi2018accurate}
\begin{figure}
    \includegraphics[width=0.95\columnwidth]{figures/fig8_sampl_scheme/Figure.pdf}
    \caption{\textbf{Four sampling strategies available in different software packages.} \textbf{A}: Multiple replicas in parallel at different lambda states. \textbf{B}: Hamiltonian replica exchange scheme. \textbf{C}: Single replica scheme sampling from all $\lambda$ states. \textbf{D}: Non-equilibrium sampling scheme.}
    \label{fig:fig_sampling_scheme}
\end{figure} 
The best practice recommendation, is to use replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme}(B)), or in absence of their availability and when looking at systems with fast mixing times parallel replica schemes are acceptable (Fig.~\ref{fig:fig_sampling_scheme}(A)). Single replica schemes and non-equilibrium schemes are not as established yet, but are very promising.

>>>>

Alchemical free energy calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor or the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}. 

@@

Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with conventional computational methods, but are more readily tractable with alchemical methods. 

@@

In alchemical simulations, the introduction of intermediate \textit{alchemical states} that bridge the high-probability regions of configuration space between the two physical endstates of interest permits the free energy of large transformations to be robustly computed.

@@

alchemical calculations can be used in a variety of scenarios, such as computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} (A)); 

@@

alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (D))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (E))~\cite{mobley2007predicting,aldeghi2016accurate,aldeghi2017predictions}; and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (F))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.

Once an alchemical simulation as been performed, the data must be analyzed to estimate the free energy for the transformation of interest.

@@

As the field of molecular simulation moves towards millisecond timescales, accurate alchemical calculations on even more challenging problems will become reasonable to perform. In the meantime, today's users may find it hard to get started with these complex calculations whilst also keeping up with the fast pace of change. This best practice guide aims to serve as a go-to reference for current recommendations and tips for users of all experience.  

@@

We assume the reader possesses a basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association. If you feel unfamiliar with some of these concepts please good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}. Oftentimes, docking calculations are required to generated initial binding pose for a small molecule for an alchemical calculation, where binding poses are not known, thus some basic familiarty on how to run docking calculations is also expected~\cite{grinter2014challenges}. 

@@

In the remainder of this paper, we will cover topics that are key to the preparation~(\ref{sec:step1}), choice and use of correct protocols~(\ref{sec:step2}), and finally the best practices that should be used for the analysis of alchemical calculations~(\ref{sec:step3}). 

@@

@@

Many practitioners expect alchemical methods to provide valuable guidance for drug discovery, and to exhibit accuracy superior to most  alternative approaches for suitable targets~\cite{kuhn2017prospective}. It has been stressed that successful application in industry may require considerable knowledge of the "domain of applicability" of free energy calculations -- where they work well and where they will not ~\cite{sherborne2016collaborating}. Successful application also requires robust protocols for preparing, submitting and analysing alchemical calculations. In this regard, the issues mentioned in the previous section such as understanding the suitability and timescales to capture the structure activity relationships (SAR), and performing up-front tests of performance are all relevant to drug discovery applications. Without venturing too far into details of system setup, which is beyond the scope of this article, we highlight some critical factors affecting accuracy and successful application. 

@@

    \caption{\textbf{Two common topologies for alchemical calculations} left: A single topology uses softcore potentials to convert from one type of atom to an other. Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms. right: The dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses softcore potentials for this. The 'mixed' intermediate atoms are used in both dual and single topology approaches. Only the way the transformation occurs and the end states differ. Following the arrow along the left and right illustrate the differences.  Figure adapted from \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}}

@@

\subsubsection{What different sampling schemes exist and which one should I use?}
With all alchemical simulations having the requirement of sampling from different sets of $\lambda$ states in common different approaches have been designed. Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes used. The simplest approach is defining a set of fixed $\lambda$ values and then run independent simulations at each of the intermediate $\lambda$ values, see Fig.~\ref{fig:fig_sampling_scheme}(A). This type of scheme is currently used for AMBER TI calculations~\cite{song2019} and for BioSimSpace~\cite{hedges2019biosimspace}. A slightly more complex approach is allowing replica mixing, meaning that each independent $\lambda$ simulation is attempted to be exchanged with a neighbouring $\lambda$ replica. This is based on ideas of Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{wendsen1986replica}. Using a Metropolis Hastings based acceptance criterion for exchanges means that the generated ensemble of all replicas is still sampling from the Boltzmann distribution and has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensionala,sugita1999replicaexchangea, woods2003developmenta, jiang2010free}. The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}(B). It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}. A third approach borrows ideas from simulated tempering~\cite{marinari1992simulateda}. In this scheme a single replica rapidly explores all of $\lambda$ space by optimally working out weights that allow switching between different intermediate $\lambda$ values, as seen in Fig.~\ref{fig:fig_sampling_scheme}(C) and also referred to as self adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in~\cite{andrearizzi2019choderalab}. The last approach available in conjunction with GROMACS makes use of non-equilibrium simulations~\cite{aldeghi2018accurate}. In this approach, only end state $\lambda$ replicas ($\lambda$=0, $\lambda=1$) are simulated at equilibrium intermediate information is generated from non-equilibrium simulations. A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme}. 
\ref{fig:fig_sampling_scheme} A:\cite{hedges2019biosimspace, song2019using } B\cite{aldeghi2015accurate, wang2015accurate,jiang2019computing} C:\cite{li2007simulated, tan2017optimally} D:\cite{aldeghi2018accurate}
\begin{figure}
    \includegraphics[width=0.95\columnwidth]{figures/fig8_sampl_scheme/Figure.pdf}
    \caption{\textbf{Four sampling strategies available in different software packages.} \textbf{A}: Multiple replicas in parallel at different lambda states. \textbf{B}: Hamiltonian replica exchange scheme. \textbf{C}: Single replica scheme sampling from all $\lambda$ states. \textbf{D}: Non-equilibrium sampling scheme.}
    \label{fig:fig_sampling_scheme}
\end{figure} 
The best practice recommendation, is to use replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme}(B)), or in absence of their availability and when looking at systems with fast mixing times parallel replica schemes are acceptable (Fig.~\ref{fig:fig_sampling_scheme}(A)). Single replica schemes and non-equilibrium schemes are not as established yet, but are very promising.
